A few stickleback suffice to transport adaptive alleles to new lakes

### 2019-04-24 Stickleback_Paper.tex
latest paper updates

Threespine stickleback provide a striking example of local adaptation to divergent habitats while still being contected by recurrent gene flow. These small fish occurin marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases smaller freshwater populations have been established ``de novo'' from marine colonists. Previous results show that a similar freshwater phenotype often evolves in independent populations soon after colonization using largely the same standing genetic variation, even in very distant geographic regions. Here we use genealogical simulations to determine the levels of gene flow that best matches observed patterns of allele sharing among habitats in stickleback, and to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and reuse of standing genetic variation. We find that rapid, repeated adaptation using a shared set of alleles that are maintained at low frequency by migration-selection balance in ocean populations occurs over a realistic range of intermediate rates of gene flow that match previous empirical population genomic studies in stickleback. Using these realistic simulations we discovered that low gene flow leads to slow, independent adaption to district habitats, whereas high gene flow leads to significant mutation load. A consequence is that population genomic scans for locally adapted alleles are more likely to succeed with higher rates of gene flow. Lastly, we find that the origin of many freshwater adapted alleles in the introduced lakes is the original generation of marine individuals that colonized the lake. Thuses simulations provide deeper context for existing studies of stickleback evolutionary genomics, and provide guidance for future empirical studies in this model. More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

The sequential fixation of new mutations has long been the canonical model for the genetics of adaptation (\textbf{cite}). While it has proven valid in numerous studies in the field as well as in the lab, this model is now rightfully understood as incomplete for many species in nature that have more complicated population structures. What role does genomic variation play in meta-populations with connection to an array of selective pressures? How could migration-selection balance and ecological speciation actually benefit species as a whole? A growing number of studies have identified instances of convergent or shared adaptive evolution in small demes that utilize standing genetic variation (SGV) found in the static population from which they split \citet{nelson2017ancient,Schrider2017,Barrett2008}. However, it is still widely unknown which variations of evolutionary processes such as gene flow, recombination, selection and mutation, promote the maintenance and re-use of such genetic polymorphisms. In addition, we would like to know how complex geographic systems, with a variety of selective pressures, interact with these evolutionary forces to help or hinder the use of SGV in newly colonized subpopulations

%\plr{I like the spirit of the first paragraph, but want to totally rewrite it}

Recently, a large flood of data resulting from evolutionary systems of organisms have come from population genomic studies powered by advances in sequencing technologies. One such organism is the threespine stickleback. Many studies of species have exhibited a long standing evolutionary history of migration-selection balance between marine and freshwater populations across the globe. The rigidity of the stickleback's ability to frequently prosper in newly created freshwater environments has made this fish a good model for understanding the genetic basis of adaptive evolution. The ancestral marine form of stickleback has given rise to millions of independently derived populations in recently de-glaciated regions of the Northern Hemisphere. While geographic isolation prevents direct migration between a large majority of these patches of freshwater populations, They appear to exhibit very similar phenotypes, as well as many shared, adaptive alleles often found to be identical by decent (IBD). Impressively, independent local adaptation of marine individuals to freshwater environments has been observed to take place in tens of generations, and the adaptive alleles are found in freshwater populations that have been geographically isolated since the end of the last ice age $\approx 13,000$ years ago. For example, in 1964 the Great Alaskan Earthquake caused an uplift of Middleton island and in turn, introduced a group of freshwater ponds around the perimeter of the island. Quickly inhabited by the surrounding marine population of stickleback, \citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be parallel to freshwater stickleback that have been separated for over thousands of years. In these freshwater stickleback, the number of lateral plates are reduced and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region as observed in the large majority of freshwater demes.

But how can evolution occur at such a rapid pace? Waiting for new mutations to arise in each lake would take much longer, and the genotypes being identical by state is even more improbable. This would seem to suggest that freshwater alleles are maintained in the marine individuals allowing the accelerated selection on SGV found in marine individuals which colonize the lake. The first clear example of the global reuse of SGV was the gene \textit{eda} which has been shown to be an important regulator for the number of lateral plates. While the low lateral plate version of this gene arose millions of years ago, it is found in freshwater ponds which have formed much more recently. Contemporary population genomic studies employing genome-wide haplotype analyses has provided evidence that \textit{most} regions of the genome that distinguish marine-freshwater genetic differences share this pattern \citep{nelson2017ancient}.

\citet{schluter2009genetics} formulated the ``transporter'' hypothesis to explain these observations,
a conceptual model in which alleles benefical in freshwater populations
are maintained at migration-selection balance in the larger oceanic population
and thus available for subsequent adaptation into new freshwater habitats.
The hypothesis seems fits available data,
but a number of quantitative questions remain.
Are the actual population sizes, migration rates, and fitness differentials
consistent with this hypothesis?
How many of these differentially selected alleles are there,
and how are they arranged within the genome?
Does adaptation to a new lake rely on presence of early-generation hybrids between freshwater and marine fish,
i.e., on long haplotypes recently inherited from freshwater-dwelling stickleback?
Or, is the process
more akin to the atomization and reconstruction effected by the Star Trek technology that motivated the hypothesis' name?
More generally, 
how do the dynamics and eventual genomic organization
depend on the level of gene flow and strength of local adaptation,
both across the entire stickleback meta-population
and separately within each lake?

In this paper, we used forward simulations 
modeled on the threespine stickleback meta-population
to ask how gene flow affects the genetic and genomic architecture of local adaptation.
We know in practice that stickleback populations in new habitats can adapt in only tens of generations \citep{lescak2015evolution};
what level of between-habitat gene flow is required for such rapid adaptation?
Genome scans for between-habitat differentiation have detected a number of causal QTL;
what proportion of the loci actually underlying adaptation do we expect to be detected by such scans?
How does our power to detect causal loci depend on the amount of gene flow?

We find that only a few stickleback per generation, per lake, 
suffices to maintain sufficient genetic variation in the ocean
so that new freshwater populations can adapt from standing variation in only tens of generations.
We also find that continued gene flow of marine individuals into recently established lakes
is not important for adaptation,
and so it is only ``downstream'' gene flow that allows freshwater adaptations to be ``transported'' between lakes.
% We show that low levels of introgression also results in noise that often presents itself as $F_{st}$ peaks when scanning across the genome suggesting introgression plays a key role in inferring causal loci . 
This rapid, local adaptation can only occur over an intermediate range of levels of gene flow --
low gene flow prevents variation from being transported,
while high gene flow prevents local adaptation.

%%%%%%%%%%%%%%%%
\section*{Methods}

>>>>

Threespine stickleback provide a striking example of local adaptation to divergent habitats while still being contected by recurrent gene flow. These small fish occur in marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases smaller freshwater populations have been established ``de novo'' from marine colonists. Previous results show that a similar freshwater phenotype often evolves in independent populations soon after colonization using largely the same standing genetic variation, even in very distant geographic regions. Here we use genealogical simulations to determine the levels of gene flow that best match observed patterns of allele sharing among habitats in stickleback, and to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and re-use of standing genetic variation. In our simulations we find that repeated adaptation uses a shared set of alleles that are maintained at low frequency by migration-selection balance in ocean populations. This process occurs over a realistic range of intermediate rates of gene flow that match previous empirical population genomic studies in stickleback. Examining these simulations more deeply reveals that low gene flow leads to slow, independent adaption to district habitats, whereas high gene flow leads to significant mutation load but an increase probability of success for population genomic scans for locally adapted alleles. We also find that the original generation of marine individuals that colonized the lakes is most often the sole source of most freshwater adapted alleles. These simulations provide deeper context for existing studies of stickleback evolutionary genomics, and provide guidance for future empirical studies in this model. More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

The canonical model for the genetics of adaptation has long been the sequential fixation of new mutations (\textbf{cite}). While this model has been very useful heuristically for guiding studies in the field and lab (\textbf{cite}), it is now rightfully understood as incomplete for many species in nature that have more complicated population structures (\textbf{cite}). In particular, empirical studies have increasingly pointed to the need to more deeply incorporate standing genetic variation (SGV) into adaptation dynamics for metapopulations inhabiting an array of habitats and experiencing diverse selective pressures while still exhibiting significant gene flow (\textbf{cite}). A growing number of empirical studies have identified instances of convergent evolution using SGV \citep{nelson2017ancient,Schrider2017,Barrett2008}. However, it is still not clear how variation in evolutionary processes - such as gene flow, recombination, selection and mutation -  can promote the maintenance and re-use of SGV, particularly during invasion and adaptation to new environments (\textbf{cite}). It is similarly unclear whether variation in these evolutionary processes can determine the genetic architecture of evolving traits via SGV. While emprical studies in the field and lab are essential, the use of simulation models of specific systems can greatly advanced our understanding of the adaptation genetic dynamics using SGV.

%\plr{I like the spirit of the first paragraph, but want to totally rewrite it}
%\wac{I did a little editing but Peter can still totally rewrite. Regardless it needs to be streamlined a fair bit}

The threespine stickleback fish has provided some of the earliest data in a recent flood showing the heterogeneous nature of divergence across genomes (\textbf{cite}) and the much more extensive use of SGV than once thought (\textbf{cite}). The ancestral marine form of stickleback has given rise to millions of independently derived freshwater populations in recently de-glaciated regions around the Northern Hemisphere (\textbf{cite}). While geographic isolation prevents direct migration among freshwater populations, they often evolve similar (but not exactly the same) phenotypes (\textbf{cite}). The most recent evolutionary genomic studies on stickleback document that while adaptation to freshwater can occur over decades (\textbf{cite}), the overall dynamic of local adaptation to marine and freshwater habitats has been occurring for millions of years (\textbf{cite}). Impressively, independent local adaptation of marine individuals to freshwater environments has been observed to take place in just tens of generations (\textbf{cite}). For example, in 1964 the Great Alaskan Earthquake caused an uplift of many islands and coastal regions throughout the Gulf of Alaska (\textbf{cite}). Studies of stickleback populations on uplifted Middleton Island showed that newly created freshwater ponds were invaded by the surrounding marine population of stickleback which evolved the freshwater syndrome of phenotypes in less than 50 years (\textbf{cite}). Amazingly, nearly a quarter of the genome diverged significantly in this same time but - with very little change in the rest of the genome - in a pattern that mirrored previous studies of freshwater stickleback populations thousands of years old (\textbf{cite}).

But how can evolution occur at such a rapid pace in this and other species? Waiting for new mutations to arise in each lake or pond would take much longer than the decades since the Alaskan earthquake (\textbf{cite}). Even more improbable is having divergence cluster in such similar genomic regions across independent populations. An alternative hypothesis, supported by the most recent population genomic analyses, is that the marjority of alleles important for freshwater adaptation are maintained in the marine individuals due to recurrent gene flow from freshwater back in to marine populations. In 2009 \citet{schluter2009genetics} proposed a conceptual model they termed the ``transporter hypothesis'' to describe the process of alleles benefical in freshwater populations are maintained at migration-selection balance in the larger oceanic population and thus available for subsequent adaptation into new freshwater habitats. The first clear example of the global reuse of SGV in stickleback was the gene \textit{eda} which has been shown to be an important regulator for the number of bony lateral plates (\textbf{cite}). While the low lateral plate version of this gene arose millions of years ago, it is found in much younger freshwater ponds around the Northern Hemisphere (\textbf{cite}). More recently, genome-wide haplotype analyses have provided evidence that \textit{most} regions of the genome that distinguish marine-freshwater genetic differences share this pattern \citep{nelson2017ancient}. These data support the transporter hypothesis, but a number of questions remain. What are the actual population sizes, migration rates, and fitness differentials consistent with this hypothesis? How many differentially selected alleles exist, how many are used at any one time, and how are they arranged within the genome?

Here we use forward simulations incorporating linkage to model the stickleback metapopulation and address these questions. We ask how variation in amount of gene flow affects the genetic architecture of local adaptation to newly created freshwater ponds. Because we are documenting the entire genealogy of all alleles, we can determine the distribution and abundance of haplotypes across all in silico marine and freshwater populations to know the timing and proportion of potentially adaptive alleles that are actually utilized in each population. In addition, we can document how these adaptive alleles are distributed across the genomes, and as a consequence determine how this may affect the efficacy of genome scans for between-habitat differentiation. We find that only a few stickleback migrants per lake per generation are sufficient to maintain SGV in the ocean, allowing new freshwater populations can adapt in only tens of generations. We also find that continued gene flow of marine individuals into recently established lakes is not necessary for adaptation as the majority of alleles are present in the initial marine colonizers. Finally, we find that this rapid, local adaptation can only occur over an intermediate range of levels of gene flow. Low gene flow prevents variation from being transported, while high gene flow prevents local adaptation. Together these simulation findings enhance and extend the existing rich empirical work on stickleback evolutionary genomics.

%%%%%%%%%%%%%%%%

\section*{Methods}
